---
title: "Hopfield network bdm"
description: |
    First attempt to compute bdm on Hopfield networks simulation data.
author:
  - name: Szymon Talaga and Mikołaj Biesaga
    affiliation: The Robert Zajonc Institute for Social Studies
    affiliation_url: www.iss.uw.edu.pl/en/
date: "`r Sys.Date()`"
output: radix::radix_article
---

```{r setup_env, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.asp = 1)
```

```{r setup}
library(tidyverse)
library(readr)
library(stringr)
library(purrr)
library(reticulate)
library(magrittr)
library(zoo)
library(boot)
use_condaenv("bdm")
theme_set(theme_bw())
```

## Magnetization

```{r read_data_magnetization}
path_mags <- 'data/base-model/Mags/'
list_of_files <- list.files(path = path_mags, pattern = '.csv')

mags <- list()
for (file_name in list_of_files) {
  name <- file_name %>%
    word(start = 2,
         end = 3,
         sep = "_") %>%
    str_extract(pattern = '\\w+_\\d')
  mags[[name]] <- read_delim(file = paste0(path_mags,file_name),
                          delim = '\t',
                          escape_double = FALSE,
                          col_names = FALSE,
                          trim_ws = TRUE) %>%
    rename(timestep = X1,
           magnetization = X2) %>%
    mutate(name = name)
}
mags_binded <- mags %>%
  bind_rows()
```

```{r mags_histograms}
mags_binded %>%
  group_by(name) %>%
  ggplot(aes(x = magnetization)) +
  geom_histogram() +
  facet_wrap(~name)
```

There is no point of doing anything with magnetization of subCritical because it is almost always 1. The only difference between subCritical_1 and subCritical_2 is that the former reaches 1 after 5 steps, and the latter after 13 steps.

To compute algorithmic complexity on magnetization data we first had to bin the data. We figured that there are two reasonable ways of doing it: compute quantiles or compute equally long bins.  The current version of Szymon's package limits computing algorithmic complexity to series with maximum 9 different elements. Therefore, we computed bins only up to this number of elements.

### Equal bins

```{r even_bins}
mags_equal_bins <- mags %>%
  map_depth(1, function(table){
    table %>%
      mutate("bins9" = cut(magnetization,
                            breaks = seq(min(magnetization),
                                         max(magnetization),
                                         (max(magnetization) - min(magnetization))/9),
                            include.lowest = TRUE,
                            right = TRUE,
                            labels = c('0','1','2','3','4','5','6','7','8')),
             "bins2" = cut(magnetization,
                            breaks = seq(min(magnetization),
                                         max(magnetization),
                                         (max(magnetization) - min(magnetization))/2),
                            include.lowest = TRUE,
                            right = TRUE,
                            labels = c('0', '1')),
             "bins4" = cut(magnetization,
                            breaks = seq(min(magnetization),
                                         max(magnetization),
                                         (max(magnetization) - min(magnetization))/4),
                            include.lowest = TRUE,
                            right = TRUE,
                            labels = c('0', '1', '2', '3')),
             "bins5" = cut(magnetization,
                            breaks = seq(min(magnetization),
                                         max(magnetization),
                                         (max(magnetization) - min(magnetization))/5),
                            include.lowest = TRUE,
                            right = TRUE,
                            labels = c('0','1','2','3','4')),
             "bins6" = cut(magnetization,
                            breaks = seq(min(magnetization),
                                         max(magnetization),
                                         (max(magnetization) - min(magnetization))/6),
                            include.lowest = TRUE,
                            right = TRUE,
                            labels = c('0','1','2','3','4','5')))
  }) %>%
  bind_rows() %>%
  group_by(name) %>%
  summarise(seq9 = list(bins9),
            seq2 = list(bins2),
            seq4 = list(bins4),
            seq5 = list(bins5),
            seq6 = list(bins6))
```

```{python even_bins_cmx}
import numpy as np
import pandas as pd
import warnings
from bdm import BDMIgnore(ndim=1, shape=) as BDM

data = r.mags_equal_bins

data.seq9 = data.seq9.apply(lambda x: np.array(x, dtype = int))
data.seq2 = data.seq2.apply(lambda x: np.array(x, dtype = int))
data.seq4 = data.seq4.apply(lambda x: np.array(x, dtype = int))
data.seq5 = data.seq5.apply(lambda x: np.array(x, dtype = int))
data.seq6 = data.seq6.apply(lambda x: np.array(x, dtype = int))

warnings.simplefilter('ignore')

bdm9 = BDM(ndim=1, shape=(i,), nsymbols=9)
bdm2 = BDM(ndim=1, shape=(i,), nsymbols=2)
bdm4 = BDM(ndim=1, shape=(i,), nsymbols=4)
bdm5 = BDM(ndim=1, shape=(i,), nsymbols=5)
bdm6 = BDM(ndim=1, shape=(i,), nsymbols=6)


mags_equal_bins = pd.DataFrame({"name": data.name,
                               "cmx9": data.seq9.apply(bdm9.bdm),
                               "ent9": data.seq9.apply(bdm9.ent),
                               "cmx2": data.seq2.apply(bdm2.bdm),
                               "ent2": data.seq2.apply(bdm2.ent),
                               "cmx4": data.seq4.apply(bdm4.bdm),
                               "ent4": data.seq4.apply(bdm4.ent),
                               "cmx5": data.seq5.apply(bdm5.bdm),
                               "ent5": data.seq5.apply(bdm5.ent),
                               "cmx6": data.seq6.apply(bdm6.bdm),
                               "ent6": data.seq6.apply(bdm6.ent)})
```

### The difference between quantiles and equal bins

```{r difference_between_bins}
mags_quantiles_cmx <- py$mags_quantiles %>%
  mutate(bins = "quantiles") %>%
  unnest()
mags_equal_bins_cmx <- py$mags_equal_bins %>%
  mutate(bins = "equal_bins") %>%
  unnest()



mags_quantiles_cmx %>%
  bind_rows(mags_equal_bins_cmx) %>%
  gather(key = "klucz",
         value = "wartosc",
         ncmx9:cmx6) %>%
  mutate(klucz = paste0(klucz,"_", name)) %>%
  select(-name) %>%
  filter(!grepl(klucz, pattern = "^ncmx\\d")) %>%
  spread(bins, wartosc) %>%
  mutate(difference = equal_bins - quantiles)
  
```


