---
title: "Hopfield netwrosk bdm"
description: |
    First attempt to compute bdm on Hopfield networks simulation data.
author:
  - name: Szymon Talaga and Mikołaj Biesaga
    affiliation: The Robert Zajonc Institute for Social Studies
    affiliation_url: www.iss.uw.edu.pl/en/
date: "`r Sys.Date()`"
output: radix::radix_article
---

```{r setup_env, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.asp = 1)
```

```{r setup}
library(tidyverse)
library(readr)
library(stringr)
library(purrr)
library(reticulate)
library(magrittr)
use_condaenv("bdm")
theme_set(theme_bw())
```

## Magnitude


```{r read_data_magnitude}
path_mags <- 'data/base-model/Mags/'
list_of_files <- list.files(path = path_mags, pattern = '.csv')

mags <- list()
for (file_name in list_of_files) {
  name <- file_name %>%
    word(start = 2,
         end = 3,
         sep = "_") %>%
    str_extract(pattern = '\\w+_\\d')
  mags[[name]] <- read_delim(file = paste0(path_mags,file_name),
                          delim = '\t',
                          escape_double = FALSE,
                          col_names = FALSE,
                          trim_ws = TRUE) %>%
    rename(timestep = X1,
           magnetization = X2)
}
```

What's the difference between critical 1 and critical 2, because their histogram are clearly different?
I am just not sure if there anything to see here if the differences are so obvious.
How to bin the data, doing it by just cutting into even parts makes no sense at all. Quantiles or whatsever has more sense.

```{r histograms}
lapply(mags, function(table){hist(table$magnetization)})
```

## States

```{r read_data_states}
path_states <- 'data/base-model/States/'
list_of_files <- list.files(path = path_states, pattern = '.csv')

states <- list()
for (file_name in list_of_files) {
  name <- file_name %>%
    word(start = 2,
         end = 3,
         sep = "_") %>%
    str_extract(pattern = '\\w+_\\d')
  states[[name]] <- read_delim(file = paste0(path_states,file_name),
                          delim = '\t',
                          escape_double = FALSE,
                          col_names = FALSE,
                          trim_ws = TRUE) %>%
    rename(timestep = X1,
           instantaneous = X2,
           preferred = X3) %>%
    group_by(timestep) %>%
    mutate(id = 1:n())
}
```



```{r}
states_bdm <- lapply(states, function(table){
  table %>%
    mutate(compability = instantaneous * preferred) %>%
    gather(key = 'klucz',
           value = 'bit',
           c(instantaneous, zgodnosc, preferred)) %>%
    mutate(bit = (bit + 1)/2) %>%
    group_by(timestep, klucz) %>%
    summarise(seq = list(bit))
})

```


```{python}
import numpy as np
import pandas as pd
from bdm import BDMIgnore as BDM

bdm = BDM(ndim=1)
    
data = pd.DataFrame.from_dict(r.states_bdm, orient = 'index').rename(columns = {0: 'frames'})
result = []

for i in range(6):
   data.frames[i].seq = pd.Series(data.frames[i].seq.apply(lambda x: np.array(x, dtype = int)))
   result.append(pd.DataFrame({'timestep': data.frames[i].timestep.astype(int),
                               'cmx': data.frames[i].seq.apply(bdm.bdm),
                               'ncmx': data.frames[i].seq.apply(bdm.nbdm),
                               'ent': data.frames[i].seq.apply(bdm.ent),
                               'nent': data.frames[i].seq.apply(bdm.nent),
                               'type': data.frames[i].klucz,
                               'name': data.index[i]}))

```
Zobaczyć jak się będzie zgodność zmieniać preferred i 
```{r}
states_cmx <- py$result %>%
  bind_rows()
```
```{r chart_cmx}

states_cmx %>%
  filter(type != 'preferred') %>%
  group_by(type) %>%
  ggplot(aes(x = timestep, y = ncmx, color = name)) +
  facet_wrap(~type) +
  geom_line() +
  labs(y = "Normalized Algorithmic Complexity",
       x = "Timestep") +
  scale_color_discrete("")

states_cmx %>%
  filter(type != 'preferred') %>%
  group_by(type, name) %>%
  summarise(sd_cmx = sd(ncmx),
            sd_nent = sd(nent))
  

```
```{r chart_cmx}

states_cmx %>%
  filter(type != 'preferred') %>%
  group_by(type) %>%
  ggplot(aes(x = timestep, y = nent, color = name)) +
  facet_wrap(~type) +
  geom_line() +
  labs(y = "Normalized Entropy",
       x = "Timestep") +
  scale_color_discrete("")

```
